Metric Spaces with Expensive Distances

### 2019-04-24 paper/expensive-distances-main.tex
interim

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered (\cite{cohen2007stability}). For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
outperforms naive computation of all pairwise distances or runs slightly slower, depending on the approximation
quality.

@@

The \emph{greedy spanner}(\cite{althofer1993sparse}) is a simple algorithm to compute linear-sized spanners:
\begin{algorithmic}
\label{alg:greedy_spanner}
\Function{GreedySpanner}{$P, \eps$}
    \State {$E\gets\emptyset$}
    \State {Sort all pairwise distances of points in $P$}
    \ForAll {pairs $(p_i,p_j)$ in increasing order}
    \State {$d_{ij}\gets$ Shortest path distance in $(P,E)$}
    \If{$d_{ij}>(1+\eps)\dist(p_i,p_j)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \EndIf
    \EndFor
    \Return {$(P,E)$}
\EndFunction
\end{algorithmic}

The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
it was also shown to return the sparsest graph.  However, it clearly
has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
this means that in our cost model, the greedy spanner has the worst possible
performance.

@@

We performed two types of experiments: on points sampled from a low-dimensional Euclidean
space and on persistence diagrams with Wasserstein distance.
Clearly, our model is not meaningful in a Euclidean space, where
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

The realistic data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only used
dimension 0, since the diagrams in this dimension are larger in size.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with the \libraryname{DIPHA}
library \cite{bauer2014distributed}.

Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}} 
any advantage of our algorithm disappears. A plausible explanation 
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in the size of the greedy spanner.
\ignore{
However, the cardinality of the \grdy
spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
then even the \grdy spanner is almost half of the complete graph.
longer edges of an optimal matching play a more important role
than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
approaches bottleneck distance $\btdist$, which depends on the longest
edge only. If all diagrams come from the same dataset,
then there is a large number of pairs of diagrams whose optimal matchings
have the longest edges of approximately the same length. This fact means that we are in the case
of high doubling dimension, being close to a clique, and there is no hope
of getting a small spanner. On the other hand, computation of $\btdist$ in practice
requires significantly less time than computation of $\wsdist{q}$, and the case
of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; for a 1.5 approximation, we
reduce the number of distances computed by a factor of 10.

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
First of all, we should explain why the timings in the brute-force columns
depend on $\eps$. Wasserstein distance can either be computed exactly,
by using the Hungarian algorithm, or
it can be computed with the user-given accuracy with the auction algorithm,
as in \hera. Exact computation is very expensive \cite{kerber2017geometry}, and in our code for experiments
with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
To make the comparison absolutely fair, when we compute a \espanner,
we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
Let us emphasize that the diagrams we have are rather small, with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time, and we expect this advantage
to grow for diagrams of larger cardinality.

>>>>

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric \cite{rubner2000earth}.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
runs faster than naive computation of all pairwise distances or runs slightly slower, depending on the approximation
quality.

@@

The \emph{greedy spanner} \cite{althofer1993sparse} is a simple algorithm to compute linear-sized spanners:
\begin{algorithmic}
\label{alg:greedy_spanner}
\Function{GreedySpanner}{$P, \eps$}
    \State {$E\gets\emptyset$}
    \State {Sort all pairwise distances of points in $P$}
    \ForAll {pairs $(p_i,p_j)$ in increasing order}
    \State {$d_{ij}\gets$ Shortest path distance in $(P,E)$}
    \If{$d_{ij}>(1+\eps)\dist(p_i,p_j)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \EndIf
    \EndFor
    \Return {$(P,E)$}
\EndFunction
\end{algorithmic}

The greedy spanner is guaranteed \cite{althofer1993sparse} to return a spanner of size $O(n)$
(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
it was also shown to return the sparsest graph.  However, it
has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
this means that in our cost model, the greedy spanner has the worst possible
performance.

@@

We performed two types of experiments: on points sampled from a low-dimensional Euclidean
space and on persistence diagrams
with Wasserstein distance\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}}.
Clearly, our model is not meaningful in a Euclidean space, where
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

The realistic data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only used
dimension 0, since the diagrams in this dimension are larger in size.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with the \libraryname{DIPHA}
library \cite{bauer2014distributed}. 
Wasserstein distance between diagrams can either be computed exactly,
by using the Hungarian algorithm, or
it can be computed with the user-given accuracy with the auction algorithm.
Exact computation is very expensive \cite{kerber2017geometry}, 
and in our experiments we always compute $\wsdist{q}$ using \hera with relative error 0.01.

Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
any advantage of our algorithm disappears. A plausible explanation
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in size of the greedy spanner.
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; for a 1.5 approximation, we
reduce the number of distances computed by a factor of 10.

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
We compared our approach for a fixed
$\eps$ with the time to compute all pairwise-distance between the
persistence diagrams using \hera, using the same approximation parameter
$\eps$. Let us emphasize that the diagrams we have are rather small,
with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time,
and we expect this advantage to grow for diagrams of larger cardinality.
